---
title: "Intro to R for Healthcare Analytics"
author: "Vafa Saboori"
output: html_document
---

## Libraries (Packages) 

```{r Libraries, message=FALSE, warning=FALSE}
library(tibble)  # Modern dataframes
library(readr)  # Imports and exports spreadsheet files
library(dplyr)  # Tidy data analysis
library(DT)  # Displays HTML tables
```

## Simple arithmetic

```{r Arithmatics}
2 + 2 + 4 - 5
2 * 4 * 3
3 / 4
3 ^ 3 # The sign is called the caret symbol.
```

## Lists

```{r List of Systolic Blood Pressures (numerical)}
c(120, 120, 110, 130, 140)
```

```{r List of diseases (strings)}
c("Pneumonia", "ARDS", "Chronic bronchitis")
```

## Computer variables (objects)

```{r Systolic blood pressure}
sbp <- c(120, 120, 110, 130, 140)
sbp # calling the sbp list object
```

```{r A sequence}
patient.number <- seq(from =1,
                      to = 100,
                      by = 2)
patient.number # call the above list
```

## Addressing

```{r Addressing}
sbp[1] # First systolic Blood Pressure
sbp[1:3] # First three systolic blood pressures
sbp[c(1,3)] # First and thirs systolic blood pressures
```

## Distributions

```{r Uniform Distribution}
set.seed(123) # Forces a specific selection of values
age <- sample(18:85,
              500,
              replace = TRUE)
age # print or call all age values
hist(age)
```

```{r Standard normal distribution}
set.seed(123) # Forces a specific selection of values
before.after <- rnorm(500) # difference in Cholesterol levels before after drug
before.after # print or call all values
```

```{r Normal distribution}
set.seed(123) # Forces a specific selection of values
sbp <- round(rnorm(500,
                   mean = 120,
                   sd = 20),
             digits = 0)
sbp # print or call all values
hist(sbp)
```

```{r Chi-squared distribution}
set.seed(123) # Forces a specific selection of values
crp <- round(rchisq(500, # C-Reactive Protein (CRP)
                    df = 2),
             digits = 1)
crp # print or call all values
hist(crp)
```

```{r Random choice}
set.seed(123)
group <- sample(c("Control",
                  "Placebo"), # patient in control or placebo group
                500,
                replace = TRUE)
group # print or call all values
```

```{r Random choice with weights}
set.seed(123)
side.effects <- sample(c("No", 
                         "Yes"), # experienced side effects
                       500,
                       replace = TRUE,
                       prob = c(0.8,
                                0.2))
side.effects # print or call all values
```

## Descriptive statistics

```{r Descriptive Statistics (age)}
mean(age)
median(age)
var(age)
sd(age)
range(age)
IQR(age)
quantile(age, 0.25) # first quartile
quantile(age, 0.75) # third quartile
```

```{r Summary of age}
summary(age) # all descriptive statistics
```

```{r Sample space of group variable}
unique(group)
```

## Data visualization

### Box-and-whisker plot

```{r Box plot}
boxplot(age)
```

```{r Box plot with title, axes labels, color}
boxplot(age,
        col = "coral",
        main = "Patient age",
        xlab = "Patients",
        ylab = "Age")
```

### Histogram

```{r Histogram of before.after}
hist(before.after,
     col = "gold",
     main = "Difference in measurement before and after treatment",
     xlab = "Measurement",
     ylab = "Counts")
```

### Scatter plot

```{r Scatter plot of numerical pairs}
plot(age,
     sbp,
     col = "forestgreen",
     main = "Age vs systolic blood pressure",
     xlab = "Age",
     ylab = "Systolic blood pressure")
```

## Tibbles

### Creating a tibble

```{r Creating a tibble from the data}
my.data <- tibble(Age = age,
                  Difference = before.after,
                  CRP = crp,
                  Group = group,
                  sBP = sbp,
                  SideEffects = side.effects)
```

### Exporting a tibble

```{r Exporting the tibble as a csv file}
write_csv(my.data,
          "Data.csv") 
```

```{r Importing dataset, message=FALSE, warning=FALSE}
data <- read_csv("ProjectData.csv")
```

## Inspecting the data

```{r Data table}
datatable(data) # DT library (package) formats the data very nicely.
```

## Selection

```{r Create new tibble of control group}
control.group <- filter(data, Group == "I")
```

```{r Patients younger than 50}
younger.patients <- filter(data, Age < 50)
```

```{r Only sBP and CRP for younger patients}
younger.patients.select <- select(
                                  filter(data, Age < 50), 
                                  sBP, CRP)
```

```{r Younger patients in group I}
younger.patients.I <- filter(data, Age < 50 & Group == "I")
```

## Descriptive statistics of the new data

### Age by group

```{r Mean age by group}
data %>% 
  group_by(Group) %>% 
  summarise(mean.age = mean(Age))
```

### Side-effect count

```{r Counting side-effects}
data %>% 
  group_by(SideEffects) %>% 
  summarise(count = n())
```

### Side effects count by group

```{r Side-effects count by group }
data %>% 
  group_by(Group) %>% 
  count(SideEffects)
```

## Visualizing the new data

### Box plot of age by group

```{r Box plot of age given the group}
boxplot(Age ~ Group,
        data = data,
        col = c("deepskyblue", "orange"),
        main = "Age distribution by group",
        xlab = "Group",
        ylab = "Age",
        las = 1)
```

```{r Scatter plot of systolic BP given the age}
plot(sBP ~ Age, # Dependent Variable first, Independent Variable next
     data = data,
     col = "blue",
     main = "Age vs systolic blood pressure",
     xlab = "Age",
     ylab = "Systolic blood pressure",
     las = 1)
```

## Inferential statistics

### Student's *t* test

```{r Student t test}
t.test(sBP ~ Group, # Compares the mean of sBP between group I and II
       data)
```

### Linear regression

```{r Predicting systolic blood pressure given age}
summary(lm(sBP ~ Age,
           data = data))
```

```{r Predicting systolic blood pressure given age + CRP}
summary(lm(sBP ~ Age + CRP,
           data = data))
```
