---
title: "Introduction"
author: "Vafa Saboori"
output: html_document
---


# Introduction to RMarkdown

## Analysis of the esoph dataset in R


Adding R Code Chunks
Enclose code in three backticks and curly braces: 
     
```{r}
     data(esoph)
```

```{r}
     ?esoph # Data from a case-control study of (o)esophageal cancer in Ille-et-Vilaine, France.
```

```{r}
     str(esoph)
     summary(esoph)
     pairs(esoph)
     plot(esoph)
     hist(esoph$ncases)
     boxplot(esoph$ncases ~ esoph$agegp)
```

Running Code Chunks
Click the green "Run" button or use `Knit` to execute and render the document.

Customizing the Output
Echo and Evaluation Options:
   echo=FALSE: Runs the code but does not show the code.
   eval=FALSE: Prevents the code from being evaluated.
   include=FALSE: Runs the code but doesn’t show the code or results.
   
```{r, echo=FALSE} 
     data(esoph)
     str(esoph)
     summary(esoph)
     pairs(esoph)
     plot(esoph)
     hist(esoph$ncases)
     boxplot(esoph$ncases ~ esoph$agegp)
```


Saving and Sharing
Save the document.
Use `Knit` to render the document, usually as HTML.
Review the HTML output.
Share the `.Rmd` file with others for reproducibility.